---
title: "Taxonomic Assignment"
author: "Kimberly Ledger"
date: "2024-10-03"
output: html_document
---

cleaning up blastn taxonomic assignment of mifish metabarcoding output 
- ran blastn on sedna (blastn.sh) using ncbi database download from July 11, 2023
- -perc_identity 98 -qcov_hsp_perc 98 
- used dadasnake output (post/filtered.seq.fasta) from kenai_mifish_output_20231027

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries}
library(tidyverse)
library(tidyr)
library(stringr)
rename <- dplyr::rename
```

## load output from taxonkit 
- this is the fasta seqs from the 20241027 run on dadasnake, and blastn using a ncbi nt database accessed on 2024-07-18 with 98% seq identity and 98% query coverage   
```{r load-output-from-taxonkit}
taxonomy <-read.delim("/home/kimberly.ledger/kenai_mifish_mb/kenai_20231027/blastn_taxlineage.txt", header = FALSE, na.strings=c(""," ","NA"))

head(taxonomy)

TEMP <- taxonomy %>%
  filter(V1 == "ASV_0050")
```


## clean up the header a bit
```{r}
# use the full taxonomy rather than the seq id to collapse identical entries
tax_df <- taxonomy %>%
  filter(V4 > 100) %>% # make sure all retained matches are >100 bp
  select(-V2, -V5, -V6, -V7, -V14, -V8, -V9, -V10, -V11, -V12, -V13, -V15, -V16) %>%  #remove unnecessary columns
  group_by(V1, V17) %>% # group by the sequence key and the full taxonomy to reduce duplicate entries
  unique() %>% # doing that reduced the number of entries
  rename(qseqid=V1, perc_id=V3, length=V4, taxonomy=V17) %>% #rename headers
  filter(perc_id > 98) %>% 
  filter(!is.na(taxonomy))
```

## calcualate the number of ASV's with taxonomic assignment
```{r how-many-total-ASVs-w-taxonomy}
tax_df %>% 
  ungroup() %>%
  select(qseqid) %>%
  unique() %>%
  nrow
```

## reformat table 
```{r}
# formatting the taxonomy variables
taxon_df <- tax_df %>%
  filter(str_detect(taxonomy, ";")) %>%
  separate(taxonomy, into=c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = ";") %>% 
  filter(!grepl("sp\\.", species, ignore.case = TRUE)) %>%
  filter(!grepl("cf\\.", species, ignore.case = TRUE)) %>%
  filter(!grepl("aff\\.", species, ignore.case = TRUE)) %>%
  filter(!grepl(" x ", species, ignore.case = TRUE)) %>%         ### remove hybrids and unknown species assignments
  filter(!grepl("\\d", species))
```


## what non-fish are here? 
```{r}
not_Actinopteri <- taxon_df %>%
  filter(class != "Actinopteri")
```

not very many non-Actinopteri ASVs so just going to consolidate with the fish ID's after range filter 

## now look at the fish and figure out what taxa are in/out of our range
```{r}
to_check_range <- taxon_df %>%
  filter(class == "Actinopteri") %>%
  ungroup() %>%
  select(species, genus, family, order, class, phylum) %>%
  unique()
```

## check ranges for species using rfishbase 
```{r}
#library(remotes)
#remotes::install_github("ropensci/rfishbase")
library(rfishbase)

#one-time download of all fishbase tables... this takes a bit 
fb_import()

#first, validate species names using rfishbase synonyms
spp_df <- synonyms(to_check_range$species)

syn <- spp_df %>% 
  filter(Status == "synonym")
  
to_check_range_validated <- to_check_range %>%
  rename(synonym = species) %>%
  left_join(syn, by = "synonym") %>%
  mutate(synonym = ifelse(is.na(Species), synonym, Species)) %>%
  select(synonym:phylum) %>%
  rename(species = synonym)

#get distribution info 
spp_distribution <- distribution(to_check_range_validated$species) %>%
  select(Species, FAO) %>%
  unique()

#add column to designate if we will consider a species as "in range"- for this study, this will be NE Pacific and Arctic Ocean 
spp_distribution <- spp_distribution %>%
  mutate(in_range = ifelse(is.na(FAO), NA, "no"),
         in_range = ifelse(FAO == "Pacific, Northeast", "yes", in_range),
         in_range = ifelse(FAO == "America, North - Inland waters", "yes", in_range))

#keep just a list of spp names and yes/no/NA for "in range"  - this way we can keep track of what spp didn't have any reference information in fishbase to determine range 
spp_range <- spp_distribution %>%
  select(Species, in_range) %>%
  unique()

#how many entries do not have range info? 
range_na <- spp_range %>%
  filter(is.na(in_range))
```

When a valid name was not found, the presence of a species in the study area was checked using the GBIF database (https://www.gbif.org/).

## some species do not have range info - manually determine if these species should be considered in range
```{r}
spp_range <- spp_range %>%
  mutate(in_range = ifelse(Species == "Sebastes cheni", "no", in_range))

in_range <- spp_range %>%
  filter(in_range == "yes") 
```

## i can tell there are some species that do not belong in the dataset, so removing those manually 
```{r}
in_range <- in_range %>%
  filter(Species != "Oncorhynchus gilae") %>%
  filter(Species != "Acipenser brevirostrum") %>%  #positive control mis-id
  filter(Species != "Boreogadus saida") %>%  
  filter(Species != "Microgadus tomcod") %>%  
  filter(Species != "Clupea harengus")  
## likely more could be filtered out here - Catostomus species??? 
```


## keep only the "in range" fish species in the taxon dataframe and all none fish 
```{r}
taxon_df_reduced <- taxon_df %>%
  filter(species %in% in_range$Species) %>%
  bind_rows(not_Actinopteri)
```

## figure out what ASVs are unambiguously a single species
```{r}
single_spp_seqs <- taxon_df_reduced %>% 
  group_by(qseqid) %>%
  add_tally(name = "n_taxa") %>%
  filter(n_taxa == 1)
```

## remove the single-species seqs from the dataframe and then rank the hits by % identity for the remaining seqs
```{r}
seq_id_diff <- taxon_df_reduced %>%
  anti_join(., single_spp_seqs) %>%  ## comment out this line if there are no single species ASVs
  select(-length) %>%
  group_by(qseqid, species, genus, family, order, class, phylum, kingdom) %>%
    mutate(seq_percID = max(perc_id)) %>%
    group_by(qseqid, species, genus, family, order, class, phylum, kingdom, seq_percID) %>%
  summarise(max(seq_percID)) %>% # take just the top hit for each taxon (for each sequence)
  select(-`max(seq_percID)`) %>%
  ungroup() %>%
  group_by(qseqid) %>%
      #mutate(id_rank = rank(desc(seq_percID), ties.method = "min")) %>% # rank the taxonomic hits per sequence by % id
       mutate(top_perc = max(seq_percID)) %>% # designate the highest % id for the best taxonomic hit in each sequence (in some, but not all cases, this is 100%)   %>%
      mutate(diff = top_perc - seq_percID) %>% # calculate the difference between the % identity of the top hit and each subsequent taxonomic hit
        arrange(diff)
```

Now you have the single best entry for each species for each sequence ranked and with the difference between the first and second ranked entries calculated.

For sequences with multiple top hits, where the difference between ranked taxa = 0, you will end up defaulting to genus- or family-level ID. We'll do the same for any sequences where the difference between the first and second ranked taxa is < 0.5%.

## Figure out which differences are **> 0.5%** and eliminate those first
```{r}
# filter out any taxa that are >1% less matching identity than the top taxonomic hit for a given sequence
to_remove_low_perc_hits <- seq_id_diff %>%
  ungroup() %>%
  filter(diff > 0.5)

keepers <- seq_id_diff %>%
  anti_join(to_remove_low_perc_hits)
```

## so now determine whether ASVs should be assigned to genus, family, order, etc. 
```{r}
singletons <- keepers %>%
  select(qseqid) %>%
  tally() %>%
  filter(n == 1)

# these are the seqs that now have only a single match
singleton_df <- singletons %>%
  left_join(keepers) %>%
  select(-n) %>%
  bind_rows(single_spp_seqs) %>% # combine the single spp data
  mutate(taxonomic_level = "species") %>%
  mutate(taxon = species)

## Genus-level matches
# remove the singletons from the bigger df 
single_genus <- keepers %>%
  anti_join(singleton_df)%>% # at best, these should be genus-level matches
  group_by(qseqid, genus) %>%
  tally() %>%
  ungroup() %>%
  group_by(qseqid) %>%
  tally() %>%
  filter(n == 1) %>% # seqs that match a single genus
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "genus") %>%
  mutate(taxon = genus)
  

## Family-level matches
single_family <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  group_by(qseqid, family) %>%
  tally() %>%
  ungroup() %>%
  group_by(qseqid) %>%
  tally() %>%
  filter(n == 1) %>% # seqs that match a single family
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "family") %>%
  mutate(taxon = family)


## Order-level matches
single_order <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  anti_join(single_family) %>%
  group_by(qseqid, order) %>%
  tally() %>%
  ungroup() %>%
  group_by(qseqid) %>%
  tally() %>%
  filter(n == 1) %>% # seqs that match a single order
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "order") %>%
  mutate(taxon = order)


## Class-level matches
single_class <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  anti_join(single_family) %>%
  anti_join(single_order) %>%
  group_by(qseqid, class) %>%
  tally() %>%
  ungroup() %>%
  group_by(qseqid) %>%
  tally() %>% 
  filter(n == 1) %>% # seqs that match a single class
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "class") %>%
  mutate(taxon = class)


## Phylum-level matches
single_phylum <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  anti_join(single_family) %>%
  anti_join(single_order) %>%
  anti_join(single_class) %>%
  group_by(qseqid, phylum) %>%
  tally() %>%
  ungroup() %>%
  group_by(qseqid) %>%
  tally() %>% 
  filter(n == 1) %>% # seqs that match a single phylum
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "phylum") %>%
  mutate(taxon = phylum)

# Kingdom level matches
single_kingdom <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  anti_join(single_family) %>%
  anti_join(single_order) %>%
  anti_join(single_class) %>%
  group_by(qseqid, kingdom) %>%
  tally() %>%
  ungroup() %>%
  group_by(qseqid) %>%
  tally() %>% 
  filter(n == 1) %>% # seqs that match a single kingdom
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "kingdom") %>%
  mutate(taxon = kingdom)
```

## modify the singleton_df to include the right variable headers
```{r}
single_spp <- singleton_df %>%
  select(-perc_id, -length, -n_taxa) %>%
  mutate(taxonomic_level = "species") %>%
  mutate(taxon = species)
```

## recombine the full data set now that the appropriate level of taxonomy has been determined
```{r}
tax_df_2 <- bind_rows(single_kingdom, single_phylum, single_class, single_order, single_family, single_genus, single_spp)
```

## there are a few ASVs that have no taxonomic assignment
```{r}
# Extract the numerical part from the ASV_ID columna
asv_numbers <- tax_df_2 %>%
  mutate(numeric = as.numeric(str_extract(qseqid, "\\d+")))

# Find the missing values in the sequence
missing_asvs <- setdiff(min(asv_numbers$numeric):max(asv_numbers$numeric), asv_numbers$numeric)

# Create missing ASV names
missing_asv_names <- sprintf("ASV_%04d", missing_asvs)
missing_asv_names
```

okay, that's quite a few. let's investigate what happened to a few 

```{r}
taxon_df %>%
  filter(qseqid == "ASV_0268")
```
ASV39 - Gasterosteus nipponicus - stickleback from japan
ASV55, ASV100, ASV117, ASV119, etc... - no match
all good here. 

## check if it is worth doing a special taxonomic assignment for salmon 
```{r}
salmon <- tax_df_2 %>%
  filter(family == "Salmonidae")
```

just four ASVs are at Oncorhynchus (genus-level) because of the pacific salmon species (other are O. mykiss, O. clarkii)

since, it's quick i'll blast the salmon db:
blastn -query filtered.seqs.fasta -db /genetics/edna/custom_dbs/oncorhynchus_db -out oncorhynchus_results.txt -perc_identity 96 -qcov_hsp_perc 98 -num_threads 10 -outfmt '6 qseqid sseqid pident length mismatch gapopen qstart qend sstart send evalue bitscore sscinames staxids'

```{r} 
salmon <- read.table("/home/kimberly.ledger/kenai_mifish_mb/kenai_20231027/post/oncorhynchus_results.txt")  %>%
  filter(V4 > 100) %>% # make sure all retained matches are >100 bp
  select(-V5, -V6, -V7, -V8, -V9, -V10, -V11, -V12, -V13, -V14) %>%  #remove unnecessary columns
  group_by(V1, V2) %>% # group by the sequence key and the full taxonomy to reduce duplicate entries
  unique() %>% # doing that reduced the number of entries
  rename(qseqid=V1, species=V2, perc_id=V3, length=V4) #rename headers
```

## rank the hits by % identity
```{r}
salmon_seq_id_diff <- salmon %>%
  select(-length) %>%
  group_by(qseqid, species) %>%
    mutate(seq_percID = max(perc_id)) %>%
    group_by(qseqid, species, seq_percID) %>%
  summarise(max(seq_percID)) %>% # take just the top hit for each taxon (for each sequence)
  select(-`max(seq_percID)`) %>%
  ungroup() %>%
  group_by(qseqid) %>%
      #mutate(id_rank = rank(desc(seq_percID), ties.method = "min")) %>% # rank the taxonomic hits per sequence by % id
       mutate(top_perc = max(seq_percID)) %>% # designate the highest % id for the best taxonomic hit in each sequence (in some, but not all cases, this is 100%)   %>%
      mutate(diff = top_perc - seq_percID) %>% # calculate the difference between the % identity of the top hit and each subsequent taxonomic hit
      arrange(diff)

salmon_seq_id_diff #%>%
   #filter(species == "Oncorhynchus_keta")
```

since there are very few basepair difference among species, let's go with species within 0.5% of the top match 
```{r}
salmon_keepers <- salmon_seq_id_diff %>%
  filter(diff < 0.5)
```

## so now determine whether ASVs should be assigned to genus, family, order, etc. 
```{r}
salmon_singletons <- salmon_keepers %>%
  select(qseqid) %>%
  tally() %>%
  filter(n == 1)

# these are the seqs that now have only a single match
salmon_singleton_df <- salmon_singletons %>%
  left_join(salmon_keepers) %>%
  select(-n) %>%
  mutate(taxonomic_level = "species") %>%
  mutate(taxon = species) #%>%
  #select(qseqid, taxon, taxonomic_level)

## Genus-level matches
# remove the singletons from the bigger df 
salmon_single_genus <- salmon_keepers %>%
  anti_join(salmon_singleton_df)%>% # at best, these should be genus-level matches
  left_join(., salmon_keepers) %>%
  arrange(qseqid) %>%
  mutate(taxonomic_level = "genus") %>%
  mutate(taxon = "Oncorhynchus") #%>%
  #select(qseqid, taxon, taxonomic_level) %>%
  #unique
```

hmm - so the custom database only helped one of those pacific salmon asvs. will just update that one manually. 

```{r}
tax_df_2 <- tax_df_2 %>%
  mutate(species = ifelse(qseqid == "ASV_0107", "Oncorhynchus nerka", species),
         taxon = ifelse(qseqid == "ASV_0107", "Oncorhynchus nerka", taxon),
         taxonomic_level = ifelse(qseqid == "ASV_0107", "species", taxonomic_level)) %>%
  unique()
```

## Create output taxonomy data frames
```{r}
# and then collapse that down to just a single taxon per ASV
# can i collapse while keeping the taxonomy for the taxon....? 
tax_df_3 <- tax_df_2 %>%
  mutate(species = ifelse(taxonomic_level == "family", NA, species),
         genus = ifelse(taxonomic_level == "family", NA, genus),
         species = ifelse(taxonomic_level == "genus", NA, species)) %>%
  select(qseqid, taxon, taxonomic_level, species, genus, family, order, class, phylum, kingdom) %>%
  unique()
```

```{r}
write.csv(tax_df_2, "/home/kimberly.ledger/kenai_mifish_mb/outputs/taxonomy_uncollapsed.csv")
write.csv(tax_df_3, "/home/kimberly.ledger/kenai_mifish_mb/outputs/taxonomy_collapsed.csv")
```

